---
title: "A Visualisation of the Derivative."
author: "Toby Hayward"
date: "4/16/2022"
output: 
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


Any calculus student encounters the derivative for the first time in their 3^rd^ to 2^nd^ to last year of high school. They'll learn that the derivative of a function is the rate of change at an instant in time, or graphically, the slope tangent line to the curve plotted by a function $f(x)$. \
They'll learn that the derivative is a function which gives this number, given a point $x$, and that we can derive this equation through applying a series of rules such as the *Power Rule*, *Chain Rule*, *Quotient Rule*, or by simply remembering the corresponding derivative equation $f'(x)$ of $f(x)$ such as $\frac{d}{dx} sin(x) = cos(x)$. \
\
```{r echo=FALSE}

f1x <- tibble(x = seq(-0.6, 3.1, 0.01),
              y = -2.4 * x^5 + 3.7 * x^4 + 0.5 * x^3 + -1.9 * x^2 + 0.7 * x + 1 + 0.39 * x^6,
              yi = -2.4 * 5 * x^4 + 3.7 * 4 * x^3 + 0.5 * 3 * x^2 + -1.9 * 2 * x + 0.7 + 0.39 * 6 * x^5) %>% 
  pivot_longer(cols = c(y, yi), names_to = 'fun', values_to = 'res')

f1x %>% ggplot(aes(x = x, colour = fun)) +
  geom_line(aes(y = res)) +
  geom_hline(yintercept = 0, col = 'darkgray') +
  geom_vline(xintercept = 0, col = 'darkgray') +
  theme_minimal() +
  scale_y_continuous(breaks = seq(-10, 6)) +
  theme(axis.title.y = element_text(margin = margin(r = 20)),
        axis.title.x = element_text(margin = margin(t = 20))) +
  labs(title = "Plot of Polynomial f(x) and it's Derivative f'(x)", y = 'y',
       colour = '') +
  scale_color_manual(values = c('black', 'red'), labels = c('f(x)', "f'(x)"))
  


```


## The problem.
Unfortunately for a lot of students, they go through their education without understanding where the derivative first came from. How did we derive the *Power Rule* and what does an instantaneous rate of change mean? \
After studying further they'll eventually encounter the limit definition for a derivative, and begin to understand how reducing this little distance $dx$ to an infinitesimally small length, they can derive the true tangent line at a point on the curve, and hence it's slope. \
\
I propose the following visualisation of how this process works. This visualisation takes a function $f(x)$ and a distance $dx$ and estimates the tangent slope between two points on the curve $f(x)$ at a horizontal distance of $dx$. It then shows by reducing $dx$ how the function becomes a better and better approximation of the derivative function $f'(x)$, hence visualising the limit definition of a derivative. 

```{r thefunct}
derivative_approx <- function(a, b, dx, fun, dfun) {
  # [a, b] = domain of the function
  # dx = distance between the points x1, x2 to estimate dy.
  # fun, dfun = the functions f(x) and f'(x) respectively. 
  
  x = seq(a, b, dx)
  y = fun(x)
  
  dy = vector(mode = 'double', length(x)) #Initialising a vector to insert values of x into.
  dy[1] = dfun(a)
  
  
  for (i in 2:((b-a)/dx + 1)) {
    dy[i] = y[i] - y[i-1]
  }
  
  ans = dy/dx
  ans[1] <- dy[1]
  
  return(ans)
  #ans = sequence of estimated slopes for a given x. 
}
```

```{r vis_functs, class.source='fold-hide'}
make_function_vis <- function(a, b, fun, ylab) {
  plot <- tibble(x = seq(a, b, (b-a)/100),
                 y = fun(x)) %>% 
    ggplot(aes(x = x, y = y)) +
    geom_line() + 
    labs(title = sprintf('Plot of f(x) against x where f(x) = %s', ylab),
         y = ylab) +
    theme_minimal() +
    theme(axis.title.y = element_text(margin = margin(r = 20)),
          axis.title.x = element_text(margin = margin(t = 20))) +
    geom_hline(yintercept = 0, col = 'darkgray') +
    geom_vline(xintercept = 0, col = 'darkgray')
    
  return(plot)
}

make_derivative_vis <- function(a, b, fun, dfun, ylab) {
  base_plot <- tibble(x = seq(a, b, (b-a)/100),
                      actualdy = dfun(x)) %>% 
    ggplot(aes(x = x, y = actualdy)) +
    geom_line(col = 'red') +
    theme_minimal() +
    theme(axis.title.y = element_text(margin = margin(r = 20)),
          axis.title.x = element_text(margin = margin(t = 20))) +
    labs(title = sprintf("f'(x) and approximates against x where f(x) = %s", ylab),
         y = "f'(x)") +
    geom_hline(yintercept = 0, col = 'darkgray') +
    geom_vline(xintercept = 0, col = 'darkgray')
  
  for (dx in seq(0.1, 1, 0.1)) {
    temp_tib <- tibble(x = seq(a, b, dx),
                       y = derivative_approx(a, b, dx, fun, dfun))
    base_plot <- base_plot +
      geom_line(data = temp_tib, aes(x = x, y = y), alpha = 1 - dx)
  }
  
  return(base_plot)
}

```

```{r vis1, fig.show='hold', out.width='50%', dpi=300}
f_name = 'x^2'

f <- function(x) {
  return(x^2)
}
df <- function(x) {
  return(2 * x)
}

a = -2; b = 2

p1a <- make_function_vis(a, b, f, f_name)
p1a

p1b <- make_derivative_vis(a, b, f, df, f_name)
p1b
```

```{r vis2, fig.show='hold', out.width='50%', dpi=300}
f_name = 'x^5 + 2x^4'

f <- function(x) {
  return(x^5+2*x^4)
}
df <- function(x) {
  return(5 * x^4 + 8 * x^3)
}

a = -2; b = 1

p2a <- make_function_vis(a, b, f, f_name)
p2a

p2b <- make_derivative_vis(a, b, f, df, f_name)
p2b
```


```{r vis3, fig.show='hold', out.width='50%', dpi=300}
f_name = 'sin(x)'

f <- function(x) {
  return(sin(x))
}
df <- function(x) {
  return(cos(x))
}

a = -2; b = 8

p3a <- make_function_vis(a, b, f, f_name)
p3a

p3b <- make_derivative_vis(a, b, f, df, f_name)
p3b
```

#### Above, notice that the darker the line gets, the closer the approximation is which is proportional to the distance of $dx$. What is important here is that in the calculation to get those black lines, **at no point did we use the actual derivative function to determine the black lines***. Therefore showing the brilliance of the limit definition of a derivative. \
*The only exception was the starting point. This was a lazy decision by me to arrive at an approximating the first term of the derivative function faster, but I hope it is still obvious that the algorithm works to approximate the derivative. This can be solved by simply plotting the line and ignoring the first term. 
